Action Groupoids and Fractional Types

### 2016-12-07 pibackground.tex
figures in sec 2; will start sec 3 next
working on sec 3
sec 3 working on it
renamed a2

\begin{figure}[ht]
\[\begin{array}{rlcl}
%% \evalone{\unitepl}{&(\inl{()})} \\
\evalone{\unitepl}{&(\inr{v})} &=& v \\
\evalone{\unitipl}{&v} &=& \inr{v} \\
\evalone{\unitepr}{&(\inl{v})} &=& v \\
%% \evalone{\unitepr}{&(\inr{()})} \\
\evalone{\unitipr}{&v} &=& \inl{v} \\
\evalone{\swapp}{&(\inl{v})} &=& \inr{v} \\
\evalone{\swapp}{&(\inr{v})} &=& \inl{v} \\
\evalone{\assoclp}{&(\inl{v})} &=& \inl{(\inl{v})} \\
\evalone{\assoclp}{&(\inr{(\inl{v})})} &=& \inl{(\inr{v})} \\
\evalone{\assoclp}{&(\inr{(\inr{v})})} &=& \inr{v} \\
\evalone{\assocrp}{&(\inl{(\inl{v})})} &=& \inl{v} \\
\evalone{\assocrp}{&(\inl{(\inr{v})})} &=& \inr{(\inl{v})} \\
\evalone{\assocrp}{&(\inr{v})} &=& \inr{(\inr{v})} \\
\\
\evalone{\unitetl}{&(\unitv,v)} &=& v \\
\evalone{\unititl}{&v} &=& (\unitv,v) \\
\evalone{\unitetr}{&(v,\unitv)} &=& v \\
\evalone{\unititr}{&v} &=& (v,\unitv) \\
\evalone{\swapt}{&(v_1,v_2)} &=& (v_2,v_1) \\
\evalone{\assoclt}{&(v_1,(v_2,v_3))} &=& ((v_1,v_2),v_3) \\
\evalone{\assocrt}{&((v_1,v_2),v_3)} &=& (v_1,(v_2,v_3)) \\
\\
\evalone{\absorbr}{&(v,\_)} &=& v \\
\evalone{\absorbl}{&(\_,v)} &=& v \\
%% \evalone{\factorzl}{&()} \\
%% \evalone{\factorzr}{&()} \\
\evalone{\dist}{&(\inl{v_1},v_3)} &=& \inl{(v_1,v_3)} \\
\evalone{\dist}{&(\inr{v_2},v_3)} &=& \inr{(v_2,v_3)} \\
\evalone{\factor}{&\inl{(v_1,v_3)}} &=& (\inl{v_1},v_3) \\
\evalone{\factor}{&\inr{(v_2,v_3)}} &=& (\inr{v_2},v_3) \\
\evalone{\distl}{&(v_1,\inl{v_3})} &=& \inl{(v_1,v_3)} \\
\evalone{\distl}{&(v_2,\inr{v_3})} &=& \inr{(v_2,v_3)} \\
\evalone{\factorl}{&\inl{(v_1,v_3)}} &=& (v_1,\inl{v_3}) \\
\evalone{\factorl}{&\inr{(v_2,v_3)}} &=& (v_2,\inr{v_3}) \\
\\
\evalone{\idiso}{&v} &=& v \\
\evalone{(c_1 \odot c_2)}{&v} &=&
  \evalone{c_2}{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inl{v})} &=&
  \inl{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inr{v})} &=&
  \inr{(\evalone{c_2}{v})} \\
\evalone{(c_1 \otimes c_2)}{&(v_1,v_2)} &=&
  (\evalone{c_1}v_1, \evalone{c_2}v_2)
\end{array}\]
\caption{\label{opsem}$\Pi$ operational semantics}
\end{figure}

@@

As an example, let $\mathbb{3}$ abbreviate the type
$(\ot \oplus \ot) \oplus \ot$. There are three values of type
$\mathbb{3}$ which are $\inl{\inl{\unitv}}$, $\inl{\inr{\unitv}}$, and
$\inr{\unitv}$. There are, up to equivalence, six combinators of type
$\mathbb{3} \iso \mathbb{3}$, each representing a different
permutation of three elements, which could be written as follows:
\[\begin{array}{rcl}
a_1 &=& \idiso \\
a_2 &=& \swapp \oplus \idiso \\
a_3 &=& \assocrp \odot (\idiso \oplus \swapp) \odot \assoclp \\
a_4 &=& a_2 \odot a_3 \\
a_5 &=& a_3 \odot a_2 \\
a_6 &=& a_4 \odot a_2
\end{array}\]
Tracing the evaluation of $a_2$ on each of the possible inputs yields:
\[\begin{array}{rcl}
\evalone{(\swapp\oplus\idiso)}{\inl{\inl{\unitv}}} &=& \inl{\evalone{\swapp}{\inl{\unitv}}} \\
&=& \inl{\inr{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inl{\inr{\unitv}}} &=& \inl{\evalone{\swapp}{\inr{\unitv}}} \\
&=& \inl{\inl{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inr{\unitv}} &=& \inr{\evalone{\idiso}{\unitv}} \\
&=& \inr{\unitv}
\end{array}\]
Thus the effect of combinator $a_2$ is to swap the values
$\inl{\inl{\unitv}}$ and $\inl{\inr{\unitv}}$ leaving the value
$\inr{\unitv}$ intact. Iterating $a_2$ again is therefore equivalent
to the identity permutation, which can be verified using the
2-combinators:
\[\begin{array}{rcl}
a_2 \odot a_2 &=& (\swapp \oplus \idiso) \odot (\swapp \oplus \idiso) \\
&\isotwo& (\swapp \odot \swapp) \oplus (\idiso \odot \idiso) \\
&\isotwo& \idiso \oplus \idiso \\
&=& \idiso
\end{array}\]

@@

For our example combinators on the type $\mathbb{3}$, simple traces
using the operational semantics show the combinator $a_1$ is the
identity permutation; the combinators $a_3$ and $a_6$ swap two of the
three elements leaving the third intact; and the combinators $a_4$ and
$a_5$ rotate the three elements. We therefore have
$\mathit{order}(a_1)=1$,
$\mathit{order}(a_2)=\mathit{order}(a_3)=\mathit{order}(a_6)=2$, and
$\mathit{order}(a_4)=\mathit{order}(a_5)=3$.

@@

\begin{lemma}
  For $p : \tau \iso \tau$, $n \in \Z$, $p^{k + n} \isotwo p^n$ where
  $k = \ord{p}$.
\end{lemma}

>>>>

\begin{figure}[ht]
{\footnotesize
\[\begin{array}{cc}
\begin{array}[t]{rlcl}
%% \evalone{\unitepl}{&(\inl{()})} \\
\evalone{\unitepl}{&(\inr{v})} &=& v \\
\evalone{\unitipl}{&v} &=& \inr{v} \\
\evalone{\unitepr}{&(\inl{v})} &=& v \\
%% \evalone{\unitepr}{&(\inr{()})} \\
\evalone{\unitipr}{&v} &=& \inl{v} \\
\evalone{\swapp}{&(\inl{v})} &=& \inr{v} \\
\evalone{\swapp}{&(\inr{v})} &=& \inl{v} \\
\evalone{\assoclp}{&(\inl{v})} &=& \inl{(\inl{v})} \\
\evalone{\assoclp}{&(\inr{(\inl{v})})} &=& \inl{(\inr{v})} \\
\evalone{\assoclp}{&(\inr{(\inr{v})})} &=& \inr{v} \\
\evalone{\assocrp}{&(\inl{(\inl{v})})} &=& \inl{v} \\
\evalone{\assocrp}{&(\inl{(\inr{v})})} &=& \inr{(\inl{v})} \\
\evalone{\assocrp}{&(\inr{v})} &=& \inr{(\inr{v})}
\end{array} &
\begin{array}[t]{rlcl}
\evalone{\unitetl}{&(\unitv,v)} &=& v \\
\evalone{\unititl}{&v} &=& (\unitv,v) \\
\evalone{\unitetr}{&(v,\unitv)} &=& v \\
\evalone{\unititr}{&v} &=& (v,\unitv) \\
\evalone{\swapt}{&(v_1,v_2)} &=& (v_2,v_1) \\
\evalone{\assoclt}{&(v_1,(v_2,v_3))} &=& ((v_1,v_2),v_3) \\
\evalone{\assocrt}{&((v_1,v_2),v_3)} &=& (v_1,(v_2,v_3))
\end{array} \\
\\
\begin{array}[t]{rlcl}
\evalone{\absorbr}{&(v,\_)} &=& v \\
\evalone{\absorbl}{&(\_,v)} &=& v \\
%% \evalone{\factorzl}{&()} \\
%% \evalone{\factorzr}{&()} \\
\evalone{\dist}{&(\inl{v_1},v_3)} &=& \inl{(v_1,v_3)} \\
\evalone{\dist}{&(\inr{v_2},v_3)} &=& \inr{(v_2,v_3)} \\
\evalone{\factor}{&\inl{(v_1,v_3)}} &=& (\inl{v_1},v_3) \\
\evalone{\factor}{&\inr{(v_2,v_3)}} &=& (\inr{v_2},v_3) \\
\evalone{\distl}{&(v_1,\inl{v_3})} &=& \inl{(v_1,v_3)} \\
\evalone{\distl}{&(v_2,\inr{v_3})} &=& \inr{(v_2,v_3)} \\
\evalone{\factorl}{&\inl{(v_1,v_3)}} &=& (v_1,\inl{v_3}) \\
\evalone{\factorl}{&\inr{(v_2,v_3)}} &=& (v_2,\inr{v_3})
\end{array} &
\begin{array}[t]{rlcl}
\evalone{\idiso}{&v} &=& v \\
\evalone{(c_1 \odot c_2)}{&v} &=&
  \evalone{c_2}{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inl{v})} &=&
  \inl{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inr{v})} &=&
  \inr{(\evalone{c_2}{v})} \\
\evalone{(c_1 \otimes c_2)}{&(v_1,v_2)} &=&
  (\evalone{c_1}v_1, \evalone{c_2}v_2)
\end{array}
\end{array}\]
\caption{\label{opsem}$\Pi$ operational semantics}
}
\end{figure}

@@

As an example, let $\mathbb{3}$ abbreviate the type
$(\ot \oplus \ot) \oplus \ot$. There are three values of type $\mathbb{3}$ which
are $ll=\inl{\inl{\unitv}}$, $lr=\inl{\inr{\unitv}}$, and
$r=\inr{\unitv}$. There are, up to equivalence, six combinators of type
$\mathbb{3} \iso \mathbb{3}$, each representing a different permutation of three
elements, which could be written as follows:
\[\begin{array}{rcl}
\permone &=& \idiso \\
\permtwo &=& \swapp \oplus \idiso \\
\permthree &=& \assocrp \odot (\idiso \oplus \swapp) \odot \assoclp \\
\permfour &=& \permtwo \odot \permthree \\
\permfive &=& \permthree \odot \permtwo \\
\permsix &=& \permfour \odot \permtwo
\end{array}\]
Tracing the evaluation of $\permtwo$ on each of the possible inputs yields:
\[\begin{array}{rcl}
\evalone{(\swapp\oplus\idiso)}{\inl{\inl{\unitv}}} &=& \inl{\evalone{\swapp}{\inl{\unitv}}} \\
&=& \inl{\inr{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inl{\inr{\unitv}}} &=& \inl{\evalone{\swapp}{\inr{\unitv}}} \\
&=& \inl{\inl{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inr{\unitv}} &=& \inr{\evalone{\idiso}{\unitv}} \\
&=& \inr{\unitv}
\end{array}\]
Thus the effect of combinator $\permtwo$ is to swap the values
$\inl{\inl{\unitv}}$ and $\inl{\inr{\unitv}}$ leaving the value
$\inr{\unitv}$ intact. Iterating $\permtwo$ again is therefore equivalent
to the identity permutation, which can be verified using the
2-combinators:
\[\begin{array}{rcl}
\permtwo \odot \permtwo &=& (\swapp \oplus \idiso) \odot (\swapp \oplus \idiso) \\
&\isotwo& (\swapp \odot \swapp) \oplus (\idiso \odot \idiso) \\
&\isotwo& \idiso \oplus \idiso \\
&=& \idiso
\end{array}\]

@@

For our example combinators on the type $\mathbb{3}$, simple traces using the
operational semantics show the combinator $\permone$ is the identity
permutation; the combinators $\permthree$ and $\permsix$ swap two of the three
elements leaving the third intact; and the combinators $\permfour$ and
$\permfive$ rotate the three elements. We therefore have
$\mathit{order}(\permone)=1$,
$\mathit{order}(\permtwo)=\mathit{order}(\permthree)=\mathit{order}(\permsix)=2$,
and $\mathit{order}(\permfour)=\mathit{order}(\permfive)=3$.

@@

\begin{lemma}
\label{lem:ordertwo}
  For $p : \tau \iso \tau$, $n \in \Z$, $p^{k + n} \isotwo p^n$ where
  $k = \ord{p}$.
\end{lemma}
